Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge

### 2019-06-06 man/ecosystems/ms_ecosystems.Rmd
UPDATE TEXT

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` and `rcompanion`  packages. The level of significance was set to 0.05 and adjusted for multiple comparisons.

@@

The trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2). Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4). Trees for southern sites showed significant positive growth trend since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High). We also observed differences in growth for southern sites. CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figures 4 and S6).

From a long-term perspective, drought events produced reduction of RWI for all sites (Figure S3), particularly in 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies. Southern sites (CA-High and CA-Low) showed weaker reductions than northern site (SJ), especially for 2005 and 2012 (Figure S3). Trees from northern site also suffered a great reduction during 1945-1946 drought event.

The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4). Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1995-2000. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression events in the last 50 years.

## Resilience of greenness and radial tree-growth to drought events
At long-term perspective, the strength of the response of the resilience indices was related to drought severity. The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (1900 - 2017) (Table S3), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336, p = 0.048) (Figure 6). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438, p = 0.019) but negative for resistance (*Rt*, $r^2$ = 0.155, p = 0.205). Trees showed the highest value of tree-growth resilience for the worst drought event in our study area (1995, see Table S3).

@@

Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3). Southern populations showed lower values for Recovery (*Rc*) than northern site for greenness and tree-growth (Tables S1 and S2). In opposite, Resistance and Resilience values were higher for southern sites than for northern ones (Tables S1 and S2).

## Climate and tree-growth
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), specially for the northern population (r > 0.6). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while no significant limitations were recorded for northen site regarding the temperatures (Figure 8b and 8c). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees.

# Discussion
## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability. 
Severe drought events negatively affected to primary and secondary growth of *Q. pyrenaica* forests in the rear edge of their distribution. Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests. These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018]. In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].

Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970 [@BladeCastroDiez2010CeEppyf; @VicenteSerranoetal2017CIG]. Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

@@

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts were also reported. Our findings are in agreenment with those studies that have showed that the assumed higher vulnerability of dry edges does not necesarily hold [@Cavin2017; @Granda2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.

@@

## Site environment shapes differential sensitivity to climate and drought of rear-edge oak populations

@@

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (e.g. soil features, tree competence, etc.) [e.g. @Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017c].

@@

## Land-use legacy effects shape distribution and sensitivity to climate change of read-edge oak populations

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008]. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017; @MesaFernandez2018]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia1991].

Our study sites have been subjected to an intense exploitation of forest resources. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015]. For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings. Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].

@@

The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). We revised the forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest], but we did not find any clearing or cutting event during 1995 - 2000 period. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017]. On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].

The only positive release events showed for CA-High site could be related with converiosn from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013; @2015VilaTraver].

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013; @Ameztegui2016] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species. Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandonded croplands and an increase in the tree-cover [@CamachoOlmedo2002].

# Conclussions

Using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively, we analyzed forest resilience to severe droughts of *Q. pyrenaica* forests located at rear edge of their distribution. Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests. However, despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q. pyrenaica* in Sierra Nevada mountain range. Surprisingly we observed high values of resilience to drought, particularly at long-term scales. This results are consistent with research showing high genetic resilience values [@Valbuena2013; @Valbuena2017], and overall suggest this mountain region still acting as refuge for deciduous *Quercus* to climate changes [@Brewer2002; @Olalde2002; @RodriguezSanchez2010].

The differences observed in tree-growth and resilience to drought between very close sites seem to confirm that responses to drought are site-dependent [@Babst2013; Cavin2017; @Dorado2017c]. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia [@VilaCabrera2019].

@@

We would like to thank Sammy Lucy Behle, M. Suarez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. We also are very thankful to F.J. Bonet-García for its valuable comments in an earlier version of the manuscript. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.

>>>>

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons.

@@

The trees of the southern population were older than those from the northern one. In addition trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4). Both growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6). Competition was similar among all sites. Yet plot basal area was greatest in CA-High (Table 1), despite the highest individual BAI growth at that site (Figure 4). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High).

Drought events produced reduction of RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies. Southern sites (CA-High and CA-Low) showed weaker reductions than the northern site (SJ), especially for 2005 and 2012 (Figure S3). Trees from northern site also suffered a great reduction during 1945-1946 drought event (Figure 5).

`$TODO$ PASAR A Discussion` The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4).

The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940 decade (the most evindent) and the second in the period 1995-2000. These periods alternate with periods of supression for over 40 years. The two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

## Resilience to drought events at the stand and individual tree levels
The strength of the response of the resilience indices was related to drought severity. The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (1900 - 2017) (Table S3), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336, p = 0.048) (Figure 6). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438, p = 0.019) but negative for resistance (*Rt*, $r^2$ = 0.155, p = 0.205). Trees showed the highest value of tree-growth resilience for the worst drought event in our study area (1995, see Table S3).

@@

Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3). The two southern populations showed lower values for Recovery (*Rc*) than the northern site both for greenness and tree-growth (Tables S1 and S2). In contrast, Resistance and Resilience values were significantly higher for southern sites than for northern ones (Tables S1 and S2).

## Tree-growth response to climate
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), being this relationship higher in the northern population (r > 0.6 *vs.* r < 0.5). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while no significant limitations were recorded for northen site regarding the temperatures (Figure 8b and 8c). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees.

# Discussion
## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability. 
Severe drought events negatively affected to primary and secondary growth of *Q. pyrenaica* forests in the rear edge of their distribution. Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests. These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].

Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range [@GeaIzquierdo2014] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970 [@VicenteSerranoetal2017CIG]. Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

@@

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts were also reported. Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necesarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.

@@

## Microenvironment shapes differential sensitivity to climate and drought of rear-edge oaks

@@

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017]

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (e.g. soil features, tree competence, etc.) [e.g. @Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b].

@@

## Land-use legacies shape distribution and sensitivity to climate change at the read-edge

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008]. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017; @MesaFernandez2018]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].

Our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before that date. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].

@@

The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). We revised the forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest], but we did not find any clearing or cutting event during 1995 - 2000 period. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004]. On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].

The only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013; @2015VilaTraver].

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Navarro2013; @DoblasMiranda2017; @Penuelas2017] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species. Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandonded croplands and an increase in the tree-cover [@CamachoOlmedo2002].

# Conclussions 
`$TODO$` las conclusiones

Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests. However, despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q. pyrenaica* in Sierra Nevada mountain range. Surprisingly we observed high values of resilience to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia [@VilaCabrera2019].

@@

We would like to thank Sammy Lucy Behle, M. Suarez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. We also are very thankful to F.J. Bonet-García for his valuable comments in an earlier version of the manuscript. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.
